\section{Conclusion}

This report now supports a clearer cross-task interim pattern: representation quality matters more than model family alone. When pretrained contextual or generative models are available, they are the strongest finished systems in every mature comparison or they at least define the strongest current baseline. At the same time, the results do not support a simple classical-versus-neural narrative, because lightweight classical baselines remain competitive and randomly initialized neural sequence models do not consistently outperform them.

For Question 1, DistilBERT is the best sentiment classifier on the matched IMDb comparison, but the gap over TF-IDF + SVM is modest relative to the much larger gap over the supervised-only BiLSTM. This indicates that the main gain comes from contextual pretraining rather than from using a neural architecture in the abstract. For Question 2, BERT again dominates the comparison and achieves the strongest overall and per-entity F1, while the feature CRF remains a strong non-transformer baseline and the BiLSTM-CRF underperforms both alternatives. For Question 3, the pretrained DistilBART summarizer outperforms TextRank on every exported lexical metric in the current capped comparison, suggesting that pretrained abstractive generation is more effective than purely extractive ranking for CNN/DailyMail-style summaries. For Question 4, the pretrained \texttt{Helsinki-NLP/opus-mt-en-de} system already provides a strong translation baseline on Multi30k, but the report does not yet contain the planned seq2seq-plus-attention comparator, so the current Q4 claim is about baseline strength rather than final relative ranking.

Question 5 is now stronger than the earlier baseline-only state because the report includes a matched trigram-versus-LSTM comparison. The trigram model remains informative as a classical reference, but the LSTM reduces test perplexity from 1388.72 to 279.46 and produces longer, more structurally sustained generations, making it the strongest finished Q5 model so far. Across Questions 1 through 5, the most defensible project-level claim is therefore that pretrained models provide the strongest empirical results when direct comparisons exist, classical baselines remain essential reference points, and supervised neural models without strong pretraining or further tuning are not reliably superior by default.

The report is still provisional in a narrower sense. Question 4 still lacks the planned seq2seq comparison, and Questions 3 and 5 may yet be extended with larger-budget or transformer-style follow-up runs. Even so, the current written sections already support the central comparative lesson of the project: the main performance gains come from representation quality and pretraining, not from adopting a neural architecture by itself.