\section{Introduction}

\subsection{Scope and Objectives}

This take-home report studies five natural language understanding and generation problems that cover both analysis and generation settings: sentiment classification, named entity recognition, summarization, machine translation, and language modeling. The common objective across all questions is not only to obtain strong final metrics, but also to compare different representation regimes in a controlled way. Wherever possible, the experiments contrast classical feature-based baselines, supervised neural architectures, and transformer or pretrained contextual models so that the final discussion can separate the effect of model class from the effect of representation quality.

At the time of this draft, all five question sections now have written report content, but they are not equally mature. Questions 1 and 2 remain the most complete direct comparisons, Questions 3 and 5 now have stable capped comparison artifacts and drafted sections, and Question 4 now has a direct transformer-versus-seq2seq comparison with an explicitly noted train-budget mismatch because the transformer is a pretrained reference. This staged approach keeps the written analysis aligned with verified artifacts rather than speculative intermediate results.

\subsection{Technical Setup}

All experiments are managed from YAML configuration files, use a fixed random seed of 42 for reproducibility, and write timestamped run directories that include the resolved configuration, package environment snapshot, metrics, and analysis artifacts. The implementation stack combines Hugging Face Datasets for dataset access, scikit-learn and related linear modeling tools for classical baselines, PyTorch for custom neural models, and transformers for pretrained contextual encoders. Report assembly follows the same principle of artifact stability: figures copied into \texttt{report/figures/} and tables placed in \texttt{report/tables/} are sourced from completed experiment outputs rather than ad hoc terminal observations.

Table~\ref{tab:introduction_task_overview} summarizes the five questions, their datasets, and the representative model families covered by the report. Some sections already contain finished direct comparisons, while others currently document strong capped or baseline-first reference points that can be extended later.

\input{tables/introduction_task_overview}

\subsection{Report Organization}

Section 2 presents Question 1, which analyzes representation learning for IMDb sentiment classification and compares TF-IDF baselines, a word-level BiLSTM, and DistilBERT under a matched larger-budget setup. Section 3 presents Question 2, which studies named entity recognition on CoNLL-2003 with CRF, BiLSTM-CRF, and BERT token classification. Section 4 presents Question 3, which now includes a direct capped TextRank-versus-DistilBART summarization comparison. Section 5 presents Question 4, which now compares a pretrained transformer reference against a custom seq2seq-plus-attention baseline on Multi30k while making the train-budget mismatch explicit. Section 6 presents Question 5, which now compares trigram, LSTM, and GPT-style transformer language-model references on a matched capped WikiText-2 setup. The conclusion section synthesizes the current cross-task findings while remaining provisional where later extensions could still change the strength of the final comparative claims.