\section{Question 3: Text Summarization}

\subsection{Dataset Description}

Question 3 compares extractive and abstractive summarization on CNN/DailyMail, a news summarization benchmark pairing long-form news articles with short highlight summaries written by editors. The architecture notes propose a 10{,}000/1{,}000/1{,}000 subset for efficient experimentation, but the currently stable head-to-head comparison is a capped run with 20 validation and 20 test examples so that both models can be evaluated on the same articles under the same export contract. The resulting numbers should therefore be read as a matched pilot comparison rather than a final large-budget benchmark.

Even with that limitation, CNN/DailyMail remains an appropriate benchmark for this comparison because the articles are substantially longer than the references and require both content selection and compression. This makes the task a natural contrast between a sentence-ranking baseline such as TextRank and a pretrained abstractive model that can rewrite source content into shorter, reference-style summaries.

\subsection{Preprocessing and Method}

The current Q3 pipeline performs shared article cleanup, truncates long inputs to 900 words, and evaluates all models with the same lexical metrics. The extractive baseline is a TF-IDF based TextRank summarizer: sentences are treated as graph nodes, pairwise cosine similarities define weighted edges, and PageRank scores determine which three sentences are selected for the final summary. The configuration uses damping 0.85 and allows up to 40 candidate sentences per article.

The abstractive baseline uses the pretrained DistilBART checkpoint `sshleifer/distilbart-cnn-12-6` without additional fine-tuning. Inputs are tokenized up to 512 tokens and decoded with beam search using output lengths between 20 and 80 tokens. This design keeps the comparison practical on the current environment while still testing the main assignment question: whether an abstractive model can produce more reference-like summaries than a purely extractive sentence selector.

\subsection{Results}

Table~\ref{tab:q3_overall_results} summarizes the matched two-model comparison. On the test split, DistilBART reaches ROUGE-1 0.4218, ROUGE-2 0.1948, ROUGE-L 0.2833, BLEU 0.1355, and METEOR 0.4254, while TextRank reaches 0.3095, 0.1042, 0.2045, 0.0674, and 0.3041 respectively. The abstractive model outperforms TextRank on every exported metric on both validation and test.

\input{tables/q3_overall_results}

The largest absolute test improvements are in METEOR (+0.1213), ROUGE-1 (+0.1124), and ROUGE-2 (+0.0906). This pattern suggests that the abstractive model improves both content selection and phrase-level overlap, not only unigram recall. Because the split is still small, these numbers should not be over-generalized, but the consistency across both splits gives a materially stronger result than the earlier baseline-only Q3 section.

\subsection{Qualitative Analysis}

The qualitative examples mirror the aggregate scores. In the teen-driver distraction article from the validation split, DistilBART produces a compact summary that preserves the central finding that distraction contributes to nearly 60\% of serious teen crashes and correctly identifies passengers and cellphones as the main sources. The corresponding TextRank output is longer, more repetitive, and much less reference-like. A second validation example on Aaron Cook's Moldovan citizenship decision shows the same pattern: both models remain on topic, but the abstractive summary keeps the citizenship, Rio 2016, and Team GB exclusion details while dropping more of the surrounding narrative.

The test split also shows that the abstractive model is not uniformly strong. In the Syrian refugee first-person article, DistilBART stays broadly on topic but still misses key reference details such as the speaker's identity and the specific 12-day journey framing. That limitation matters because it shows that better lexical overlap does not automatically guarantee perfect factual focus. Even so, across the inspected examples DistilBART is generally more concise and closer to the structure of the human-written highlights, whereas TextRank often copies lead sentences or redundant clauses directly from the article.

\subsection{Discussion}

The current Q3 evidence now supports a clear preliminary conclusion: on this task, a pretrained abstractive baseline is stronger than the extractive TextRank baseline under the same capped evaluation protocol. TextRank remains useful as a cheap and interpretable baseline, and it can preserve source facts when important content is already concentrated in the lead sentences. However, it is weaker at compression and at matching the short editor-style summaries used as references in CNN/DailyMail.

This is still not the final word on Question 3 because the comparison is based on a 20/20 capped split and zero-shot pretrained inference rather than a larger matched benchmark or task-specific fine-tuning. Even so, the section is no longer extractive-only: there is now a stable direct comparison, and the current best Q3 baseline is the pretrained DistilBART model rather than TextRank.