\section{Question 4: Machine Translation}

\subsection{Dataset Description}

Question 4 uses the English-to-German Multi30k translation benchmark \cite{elliott2016multi30k}. In this project, the dataset is accessed through its Hugging Face mirror and treated as a short-caption translation task. That makes it a useful benchmark for testing whether a model can produce fluent target-language sentences while preserving compact visual descriptions from the source caption. The current Q4 evidence now includes two finished systems evaluated on the same capped 100-example validation and 100-example test splits: a pretrained transformer reference and a custom seq2seq-plus-attention baseline.

The comparison is direct at evaluation time but not perfectly budget-matched at training time. The transformer run keeps a 100-example train slice only for artifact-contract consistency and does not perform task-specific fine-tuning, whereas the seq2seq model is trained from scratch on 2000 capped training examples. This distinction matters for interpretation, so the section should be read as a practical pretrained-versus-classical comparison rather than a matched from-scratch study.

\subsection{Preprocessing and Model}

The stronger system uses a pretrained Opus-MT English-to-German checkpoint from Helsinki-NLP \cite{tiedemann2020opusmt}. It is built on a transformer-style translation stack \cite{vaswani2017attention}. Source and target casing are preserved, inputs and decoded outputs are capped at 96 tokens, and generation uses beam search with 4 beams and length penalty 1.0. Instead of fine-tuning on Multi30k, this pipeline evaluates the checkpoint directly as a strong practical reference.

The classical comparator is a compact GRU encoder-decoder with additive attention \cite{bahdanau2015nmt}. It uses 128-dimensional token embeddings and hidden states, a single recurrent layer, teacher forcing during training, and a maximum decoded length of 32 tokens. Unlike the transformer reference, this seq2seq model is trained from scratch on a lowercased 2000-example training split. It must therefore learn translation behavior directly from the capped Multi30k data.

Evaluation uses BLEU and ChrF on both validation and test. BLEU gives a familiar n-gram precision view of translation quality, while ChrF is more tolerant of inflectional and morphological variation and is often informative for German output. Together, the two metrics provide a lightweight but credible baseline suite for the current stage of Q4.

\subsection{Results}

Table~\ref{tab:q4_overall_results} summarizes the finished comparison. The pretrained transformer reaches validation/test BLEU 0.4008/0.3572 and validation/test ChrF 0.6569/0.6380, while the stronger seq2seq reference run reaches 0.1284/0.1348 BLEU and 0.3491/0.3548 ChrF. On the shared validation/test evaluation, the transformer therefore leads the seq2seq model by 0.2724/0.2224 BLEU and 0.3078/0.2832 ChrF.

\input{tables/q4_overall_results}

The size of this gap still makes the main takeaway straightforward: under the current capped setup, the from-scratch seq2seq baseline improves materially with more training budget but is still not competitive with a strong pretrained translation checkpoint. At the same time, the comparison should not be oversold as a perfectly aligned training-budget experiment, because the transformer benefits from large-scale pretraining rather than the capped Multi30k train split.

\subsection{Qualitative Analysis}

The qualitative examples reinforce the metric gap. On test example \texttt{test\_63}, the transformer exactly matches the reference translation for \textit{Two men playing guitar in front of a large audience.}, which is the ideal behavior for a short caption translation task. Even its weaker exported example, \texttt{test\_8}, remains understandable: it rewrites the running-man caption as a longer relative-clause construction that is less concise and less reference-like than the gold translation, but it still preserves most of the core meaning.

The seq2seq outputs show why the classical baseline still remains far behind despite the stronger run. Its best exported test example, \texttt{test\_48}, captures much of the source meaning in \textit{ein älterer mann spielt saxophon. ein videospiel.}, but it still introduces an incorrect content word and awkward sentence structure. Its weakest exported example, \texttt{test\_94}, drifts badly into \textit{drei chefköche, darunter auf dem rücken zu zu}, which is grammatical only in fragments and no longer faithful to the source sports caption. In other words, the seq2seq model now recovers more partial meaning than the weaker earlier run, but it still fails to sustain fluent, source-faithful translation consistently.

\subsection{Discussion}

The current Q4 evidence now supports a clearer conclusion than the earlier baseline-only draft. The pretrained transformer is the strongest finished translation system by a wide margin and should be treated as the main Q4 result, while the seq2seq model serves as the classical reference point that shows how difficult translation remains for small from-scratch models on limited data.

The remaining caveat is experimental alignment. Because the transformer is evaluated as a pretrained system rather than fine-tuned on Multi30k, a more budget-aligned next step would be either stronger seq2seq tuning or task-specific transformer fine-tuning on a shared train split. For the current report state, however, the direct comparison already answers the assignment question well enough to show that pretraining dominates the classical seq2seq baseline even after the stronger 2000-example seq2seq run is used.