\section{Question 5: Language Modeling}

\subsection{Dataset Description}

Question 5 uses a capped WikiText-2 setup drawn from the raw \texttt{wikitext-2-raw-v1} split \cite{merity2016wikitext}. The refreshed comparison evaluates 3{,}000 training lines, 400 validation lines, and 400 test lines, which correspond to 258{,}079, 32{,}527, and 35{,}059 tokens after preprocessing. This budget is still much smaller than a full language-model benchmark, so the reported perplexities should be read as a controlled comparison across a classical n-gram baseline, a compact recurrent neural baseline, and a practical pretrained transformer reference rather than as final large-budget results.

WikiText-2 remains appropriate because it contains longer-form Wikipedia prose rather than isolated short sentences. That makes it useful for testing both local collocations and longer-range continuation behavior. A later smaller LSTM rerun also exists elsewhere in \texttt{outputs/}, but the report section below intentionally focuses on the matched 3{,}000/400/400 held-out split so that the three finished Q5 models can be compared on the same validation and test data.

\subsection{Model and Preprocessing}

The data-trained Q5 baselines use the same lowercase word-level tokenization, the same minimum token-frequency threshold of 2, and the same \texttt{<unk>} handling so that validation and test perplexity are measured against a consistent vocabulary. The classical baseline is a trigram language model with add-k smoothing and \(\alpha = 0.1\), which provides a transparent count-based reference point for next-token prediction.

The second comparator is a compact two-layer word-level LSTM \cite{hochreiter1997lstm} with 128-dimensional embeddings and hidden states, dropout 0.2, tied input-output weights, and teacher-forced next-token training. This is still a modest recurrent model rather than a state-of-the-art language model, but it is sufficient to test whether a learned hidden-state representation materially improves perplexity and generated-text fluency over the trigram baseline under the same capped dataset budget.

The third model is a practical GPT-style reference using pretrained \texttt{distilgpt2}. Unlike the trigram and LSTM models, this transformer is not trained from scratch on the capped Q5 split; instead, it is evaluated directly on the same held-out WikiText-2 lines and prompted with the same seeded generation prefixes. That makes it a useful reference point for how much additional modeling strength a pretrained causal transformer contributes beyond the data-limited baselines.

\subsection{Results}

Table~\ref{tab:q5_overall_results} reports the refreshed Q5 perplexity comparison. The trigram model reaches validation/test perplexity 1392.54/1388.72, the LSTM reduces those values to 286.06/279.46, and the pretrained distilGPT-2 reference further lowers them to 109.78/106.44.

\input{tables/q5_overall_results}

Two comparisons matter here. First, the LSTM cuts validation perplexity by 79.5\% and test perplexity by 79.9\% relative to the trigram baseline, which shows that even a compact recurrent language model is far better than sparse count-based context matching on held-out WikiText-2 continuations. Second, distilGPT-2 lowers validation perplexity by a further 61.6\% and test perplexity by 61.9\% relative to the matched LSTM result, making the pretrained transformer the strongest finished Q5 model.

\subsection{Generated Text Samples}

The exported generations are consistent with the perplexity comparison. The trigram model can maintain short-range lexical continuity for simple prompts, but it often stops early or drifts into topic fragments that are only locally plausible. Starting from the seed \texttt{the}, it generates the short fragment \textit{the support of the season started .}, which is locally acceptable but not coherent in a broader discourse sense.

The LSTM continuations are still noisy, but they are longer and generally more structurally sustained under the same seed prompts. Its generated samples for \texttt{the}, \texttt{in the}, and \texttt{he was} continue for roughly 26--27 tokens, compared with a trigram average of 17.7 tokens across the same prompts. The LSTM still exhibits repetition and topic drift, yet it preserves local grammatical structure for longer spans than the sparse n-gram model.

The distilGPT-2 generations are slightly shorter on average than the LSTM samples (24.3 vs. 26.7 tokens), but they are noticeably smoother locally. For example, the \texttt{he was} prompt continues as \textit{he was a little bit too late and it was time to try new things with my husband.}, which is still imperfect and not guaranteed to remain on-topic, yet it is substantially more fluent than the corresponding LSTM and trigram outputs.

\subsection{Discussion}

The finished Q5 comparison now makes the progression from classical to neural to pretrained-transformer modeling explicit. The trigram model remains valuable as a classical reference, but its high perplexity and unstable generations show that short fixed-context counts are not enough to model Wikipedia-style prose well. The LSTM demonstrates that even a compact recurrent model can reuse hidden-state context to reduce perplexity dramatically.

For the current report state, the trigram run should be kept as the classical baseline, the LSTM should be presented as the compact neural comparator, and distilGPT-2 should be presented as the strongest finished Q5 reference. If Q5 is reopened later, the most meaningful follow-up is a larger matched rerun or actual transformer fine-tuning rather than additional n-gram work.